---
title: "Boston Public Schools"
author: "Derek Sollberger"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
 html_document:
   toc: TRUE
   theme: cerulean
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
library("corrplot")
library("readxl")
library("tidyverse")
```

# Loading the Data

Since we are working with an Excel spreadsheet, we will use the `read_xlsx()` command.

```{r, message = FALSE, warning = FALSE}
df <- read_xlsx("BPS_air_quality.xlsx")

df <- df |>
  filter(Value >= 0) #removing some strange outliers (sensors were broken?)
```

The information is now stored in a *data frame* called `df`.

# Exploring the Data

We can use the `head()` command to get a quick look at the data.

```{r}
head(df)
```
The `str()` gives us the structure of the data frame.  Here, we can see which variables are numerical and which are categorical.

```{r}
str(df)
```
Source: https://www.bostonpublicschools.org/Page/8810

"Indoor Air Quality sensors have been installed in all BPS classrooms, Nurse’s Offices, and Main Offices, with a roof-top unit installed in every school building to measure outdoor air as a baseline per location.

"Each sensor will record the following IAQ measures:

*    Carbon Dioxide (CO2, ppm) 
*    Carbon Monoxide (CO, ppm)
*    Airborne particulates (PM10, ug/m3)
*    Airborne Particulates (PM2.5, ug/m3)
*    Temperature (T, Fahrenheit)
*    Relative Humidity (RH, percent)

# Standard Deviations

```{r}
df |>
  group_by(Parameter) |>
  mutate(sd = sd(Value, na.rm = TRUE)) |>
  ungroup() |>
  ggplot() +
  geom_point(aes(x = Value, y = sd, color = Parameter)) +
  labs(title = "Boston Public Schools",
       subtitle = "Air Quality Data",
       caption = "Bio 18",
       x = "measurement",
       y = "standard deviation") +
  theme_minimal()
```

# Range Rule of Thumb

For now (to ease ourselves into learning the concept of the standard deviation), let us use the **range rule of thumb** interval.  It is based on classical statistical textbooks that would note how "About 95 percent of data fall within two standard deviations of the mean"
$$(\bar{x} - 2s, \bar{x} + 2s)$$

We need the sample statistics for our data

```{r}
df |>
  group_by(Parameter) |>
  summarize(xbar = mean(Value, na.rm = TRUE),
            median = median(Value, na.rm = TRUE),
            sd = sd(Value, na.rm = TRUE))
```

# Boxplots

A **boxplot** tends to be used to quickly show numerical data across a categorical variable.

* the middle line is the median (50th percentile)
* the top and bottom of the rectangles are the 75th and 25th percentile respectively
* the IQR (inter-quartile range) is the difference between the 75th and 25th percentiles
* the "whiskers" are usually drawn out 1.5*IQR beyond the rectangle
* we might consider data beyond the whiskers to be outliers

```{r}
df |>
  filter(Parameter == "Carbon Dioxide" | Parameter == "Carbon Monoxide") |>
  group_by(Parameter) |>
  summarize(min = round(min(Value, na.rm = TRUE),4),
            q25 = round(quantile(Value, probs = 0.25, na.rm = TRUE),4),
            median = round(median(Value, na.rm = TRUE),4),
            q75 = round(quantile(Value, probs = 0.75, na.rm = TRUE),4),
            max = round(max(Value, na.rm = TRUE),4))
```

```{r}
df |>
  filter(Parameter == "Carbon Dioxide") |>
  ggplot() +
  geom_boxplot(aes(x = Parameter, y= Value)) +
  labs(title = "Boston Public Schools",
       subtitle = "Air Quality Data",
       caption = "Bio 18",
       x = "pollutant",
       y = "amount (ppm)") +
  theme_minimal()
```

```{r}
df |>
  filter(Parameter == "Carbon Monoxide") |>
  ggplot() +
  geom_boxplot(aes(x = Floor, y= Value)) +
  labs(title = "Boston Public Schools",
       subtitle = "Air Quality Data",
       caption = "Bio 18",
       x = "floor",
       y = "amount (ppm)") +
  theme_minimal()
```

What is the 95th percentile of carbon monoxide levels?

```{r}
df |>
  filter(Parameter == "Carbon Dioxide" | Parameter == "Carbon Monoxide") |>
  group_by(Parameter) |>
  summarize(q05 = round(quantile(Value, probs = 0.05, na.rm = TRUE),4),
            q25 = round(quantile(Value, probs = 0.25, na.rm = TRUE),4),
            q50 = round(quantile(Value, probs = 0.50, na.rm = TRUE),4),
            q75 = round(quantile(Value, probs = 0.75, na.rm = TRUE),4),
            q95 = round(quantile(Value, probs = 0.95, na.rm = TRUE),4))
```


```{r}
df |>
  filter(Parameter == "Carbon Monoxide") |>
  filter(Value < 0.9605) |>
  ggplot() +
  geom_boxplot(aes(x = Floor, y= Value, fill = Floor), 
               color = "black") +
  labs(title = "Boston Public Schools",
       subtitle = "Air Quality Data",
       caption = "Bio 18",
       x = "floor",
       y = "amount (ppm)") +
  theme_minimal()
```

# Building a Correlation Plot

```{r}
d13_data <- df
# making easier-to-use column names
colnames(d13_data) <- c("school", "floor", "room",
                  "sensor", "status", "read_date_time",
                  "parameter", "value", "units")

# This is too heavy-handed, but I am going to today group the pollution metrics
# by school and day (and average the values?)
# oh, all of the data were from the same day!
d13_data <- d13_data |>
  select(school, parameter, value) |>
  group_by(school, parameter) |>
  mutate(avg_value = mean(value, na.rm = TRUE)) |>
  ungroup() |>
  select(school, parameter, avg_value) |>
  distinct()
head(d13_data)
```

```{r}
# here I wanted each measurement type to have its own column
d13_data_wide <- d13_data |>
  pivot_wider(names_from = parameter,
              values_from = avg_value) |>
  select(-school) #to keep only numerical columns
head(d13_data_wide)
```

```{r}
corrplot::corrplot.mixed(cor(d13_data_wide,
                             use = "pairwise.complete.obs"))
```

```{r}
corrplot::corrplot(cor(d13_data_wide, use = "pairwise.complete.obs"),
                   addCoef.col = "black",
                   col = COL2("PRGn"),
                   diag = FALSE,
                   method = "square",
                   # order = "FPC",
                   tl.col = "black",
                   tl.srt = 60, #angled labels
                   type = "lower")
```